---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(psycheval)
```

# psycheval

<!-- badges: start -->
<!-- badges: end -->

The psycheval package is a set of functions that can be useful in psychological evaluations. It accompanies the [*Individual Psychometrics*](https://individual-psychometrics.rbind.io/) online textbook.

## Installation

You can install the development version of psycheval like so:

``` r
remotes::install_github("wjschne/psycheval")
```

# Functions

## Convert a variable to standard scores

Suppose you have a scaled score of 12 (*&mu;* = 10, *&sigma;* = 3) that you want to convert to a standard score (*&mu;* = 100, *&sigma;* = 15).

```{r}
library(psycheval)
x2standard(12, mu_x = 10, sigma_x = 3)
```

To convert to a z-score (*&mu;* = 0, *&sigma;* = 1):

```{r}
x2standard(12,
           mu_x = 10, sigma_x = 3,
           mu_new = 0, sigma_new = 1)
```


To convert to T-score (*&mu;* = 50, *&sigma;* = 10):

```{r}
x2standard(12,
           mu_x = 10, sigma_x = 3,
           mu_new = 50, sigma_new = 10)
```
