---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  dev = "ragg_png",
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(psycheval)
options(digits = 4)
```

# psycheval

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The psycheval package is a set of functions that can be useful in psychological evaluations. It accompanies the [*Individual Psychometrics*](https://individual-psychometrics.rbind.io/) online textbook.

## Installation

You can install the development version of psycheval like so:

``` r
remotes::install_github("wjschne/psycheval")
```


# Convert a variable to standard scores

Suppose you have a scaled score of 12 (*&mu;* = 10, *&sigma;* = 3) that you want to convert to a standard score (*&mu;* = 100, *&sigma;* = 15).

```{r}
#| label: standardscore
library(psycheval)
x2standard(12, mu_x = 10, sigma_x = 3)
```

To convert to a z-score (*&mu;* = 0, *&sigma;* = 1):

```{r}
#| label: x2standard
x2standard(12,
           mu_x = 10, sigma_x = 3,
           mu_new = 0, sigma_new = 1)
```


To convert to T-score (*&mu;* = 50, *&sigma;* = 10):

```{r}
#| label: x2t
x2standard(12,
           mu_x = 10, sigma_x = 3,
           mu_new = 50, sigma_new = 10)
```

# Covariance of weighted sums

Let's create a covariance matrix of 5 variables that all correlate at .60.

```{r}
#| label: covariancematrix
# Create covariance matrix
Sigma <- matrix(0.6, nrow = 5, ncol = 5)
diag(Sigma) <- 1
Sigma

```

Suppose we want to create a sum of the first 2 variables and a sum of the remaining three variables. What would the covariance matrix of the summary variables be?

First we create a weight matrix `w` with 5 rows and 2 columns consisting entirely of zeros. The first column represents the weight for the first variable. We put ones in the first two rows of the first column and ones in rows 3--5 in the second column. 

```{r}
#| label: wmatrix
# Create weight matrix
w <- matrix(0, nrow = 5, ncol = 2)
w[1:2,1] <- 1
w[3:5,2] <- 1
w
```

# Composite covariance

The covariance matrix of the sums is:

```{r}
#| label: compostecov
# covariance matrix of weighted sums
composite_covariance(Sigma, w)
```

To include the original variables in the covariance matrix, append an identity matrix to `w`:

```{r}
#| label: expandedw
# A 5 by 5 idendity matrix
I <- diag(5)
I

# Prepend the identity matrix to the w matrix
w_expanded <- cbind(diag(5), w)
w_expanded

# Covariance matrix of original variables and the summary variables
BigSigma <- composite_covariance(Sigma, w_expanded)
BigSigma

# Convert to correlations with the cov2cor function
composite_covariance(Sigma, w_expanded, correlation = T)
```


# Compute a composite score

Suppose we have two subtest scores *A* = 12 and *B* = 14. Both scores are on the scaled score metric (*&mu;* = 10, *&sigma;* = 3). We want to create a composite score on the standard score metric (*&mu;* = 100, *&sigma;* = 15).

The subtests *A* and *B* correlated at *&rho;* = .6.

```{r}
#| label: compositescore
# Subtest scores
x <- c(12, 14)

# Subtest correlation
rho <- 0.6

# Subtest correlation matrix
R <- matrix(c(1,rho, rho, 1), ncol = 2)

# Compute composite score
composite_score(x = x, 
                R = R,
                mu_x = 10,
                sigma_x = 3,
                mu_composite = 100,
                sigma_composite = 15)

```


# Compute a multivariate confidence interval conditioned on a set of observed scores


```{r}
#| label: multivariateci
library(readr)
library(ggplot2)
library(dplyr)

# Observed scores
x_wisc <- c(
  vci = 130,
  vsi = 120,
  fri = 123,
  wmi = 116,
  psi = 97)

# Reliability coefficients
rxx_wisc <- c(
  vci = .92,
  vsi = .92,
  fri = .93,
  wmi = .92,
  psi = .88)

# Correlation matrix
R_wisc <- ("
  index	vci 	vsi 	fri 	wmi 	psi
  vci  	1.00	0.59	0.59	0.53	0.30
  vsi  	0.59	1.00	0.62	0.50	0.36
  fri  	0.59	0.62	1.00	0.53	0.31
  wmi  	0.53	0.50	0.53	1.00	0.36
  psi  	0.30	0.36	0.31	0.36	1.00") |> 
  read_tsv(col_types = cols(
    .default = col_double(),
    index = col_character())) |> 
  tibble::column_to_rownames("index") |>
  as.matrix()
R_wisc

d_ci <- multivariate_ci(
  x = x_wisc,
  r_xx = rxx_wisc,
  mu = rep(100, 5),
  sigma = R_wisc * 225)

d_ci 

```

Compare the multivariate CI to the univariarte CI:

```{r, echo = FALSE, fig.height=7}
#| label: fig-multivariateci
#| fig-cap: Multivariate Confidence Intervals
#| fig-alt: Comparing Univariate and Multivariate Confidence Intervals
d_ci |> 
  tidyr::pivot_longer(-variable:-r_xx) |> 
  tidyr::separate(name, c("statistic", "type")) |> 
  tidyr::pivot_wider(names_from = statistic, values_from = value) |> 
  mutate(score = toupper(x) |> forcats::fct_inorder(),
         type = stringr::str_to_title(type) |> factor(levels = c("Univariate", "Multivariate"))) |> 
  ggplot(aes(score, mu, color = type)) +
  geom_point(aes(y = x), color = "gray20", size = 2.5) +
  geom_text(aes(y = x, label = x), color = "gray30", vjust = -1) +
  geom_pointrange(aes(ymin = lower, ymax = upper), position = position_dodge(0.8)) +
  geom_vline(xintercept = 0:5 + .5, size = 0.25, color = "gray70") +
  labs(x = NULL, color = "95%\nConfidence\nInterval") +
  theme_minimal(base_size = 16) + 
  theme(panel.grid.major.x = element_blank()) + 
  scale_y_continuous("Standard Scores", 
                     breaks = seq(40, 160, 15), 
                     minor_breaks = seq(40, 160, 5), 
                     limits = c(40, 160))
```


# Relative Proficiency Index

When a typical same-age peer with an ability of W = 500 has a .90 probability of answering a question correctly, what is the probability that a person with ability of W = 520 will answer the question correctly?

```{r}
#| label: rpi
rpi(x = 520, mu = 500)
```


Flipping the previous question, when a person with ability of W = 520 has a .90 probability of answering a question correctly, what is the probability that a typical same-age peer with an ability of W = 500 will answer the question correctly?

```{r}
#| label: rpireverse
rpi(x = 520, mu = 500, reverse = TRUE)
```


Criteria other than .9 proficiency are also possible. For example, when a typical same-age peer with an ability of W = 500 has a .25 probability of answering a question correctly, what is the probability that a person with ability of W = 520 will answer the question correctly?

```{r}
#| label: grpi
rpi(x = 520, mu = 500, criterion = .25)
```
